Cluster Difference Imaging Photometric Survey. III. An Open Cluster With a 300\,pc Halo

### 2021-02-26 cluster_paper/ms.tex
update paper

%; B3V at the turnoff
As part of this broader project, we focus here on a modest question:
in just a single rich southern open cluster, is the cluster halo truly
coeval with the core?  We chose NGC\,2516 for this analysis since it
is young ($\approx$120\,Myr) and sufficiently nearby
($d\approx400$\,pc) to facilitate photometric rotation measurements
using TESS, and some degree of spectroscopic analysis.  Its halo is
decidedly non-spherical: \citetalias{kounkel_untangling_2019} reported
its dimensions as $\approx20\times10\times350$\,pc.  We want to know:
is the halo real? To what extent can we use Gaia alone to reliably
identify age-dated needles in the haystack of field stars?  What are
the implications, observationally and theoretically, if we can
identify the dispersed halos of open clusters?

@@

\section{A 300\,pc Halo around a Core?}
\label{sec:gaia}
%
% What is the core? What is the halo?
%

@@

We selected candidate \cn\ members based on those available in the
literature.  While a number of pre-Gaia catalogs were available, the
purity and accuracy of the Gaia-derived results are expected to be the
current state of the art.
We therefore adopted what we viewed as the most interesting clustering
samples to compare: those of \citet{cantatgaudin_gaia_2018}
(\citetalias{cantatgaudin_gaia_2018}) and
\citet{kounkel_untangling_2019}
(\citetalias{kounkel_untangling_2019})\footnote{A third sample was
also recently reported by \citet{meingast_2021}. Although the results
of this study were made publicly available too late to be included in
our analysis,  1577 of the 1860 \citet{meingast_2021} sources (85\%)
were included in the \citet{kounkel_untangling_2019} sample.  A
complete set of \cn\ members could very well include some of the newly
reported candidate members from \citet{meingast_2021}.}.  A useful
visualization of these samples is available online\footnote{
  \url{https://homepage.univie.ac.at/stefan.meingast/coronae.html},
  made by \citet{meingast_2021}.  }.
While we could have performed our own clustering analysis based on the
Gaia data, such an effort would be replicating the work of these
investigators. And why would our clustering methods be better than
theirs? We opt instead to use their studies as starting points.

% Theia 613
\citetalias{cantatgaudin_gaia_2018} reported 1106 candidate members of
\cn, brighter than $G=18$\,mag.  \citet{kounkel_untangling_2019}
reported 3003 candidate members, and included stars up to
$\approx$1\,mag fainter.  Each study used unsupervised clustering
based on the second Gaia Data Release, detailed in
Appendix~\ref{app:clustering}.
%FIXME 
\citetalias{cantatgaudin_gaia_2018} considered a cone of radius {\bf
XX DEGREES} centered on the cluster's position.
Given the classically reported tidal radius of {\it e.g.,} 9\,pc, CITE
Piskunov+2008, which is 8$^\circ$ at the cluster distance of
$\approx$400\,pc, perhaps they should have considered stars further
out.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
			\includegraphics[width=0.85\textwidth]{f2a.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf HR diagram of NGC\,2516.}
    No reddening correction has been applied; the photometric data are
    from Gaia EDR3.  The core (black) shows a main sequence and
    turnoff consistent with stars of a fixed age and metallicity.  The
    halo (blue) is similar, though the binary sequence is less
    well-defined.  The faintest M dwarfs in the core and halo are
    brighter than in the field star comparison sample (gray),
    consistent with these stars having not yet reached the ZAMS.
    \label{fig:hr}
  }
\end{figure*}

The first check on whether this membership assignment is plausible was
already performed by \citetalias{cantatgaudin_gaia_2018},
\citetalias{kounkel_untangling_2019}, and more recently by
\citet{meingast_2021}.  That check is to see whether the HR diagrams
of the cluster members support the claim that they are coeval.

Figure~\ref{fig:hr} presents similar results to what these
investigators have already found.  The core members of the cluster
show a clean sequence consistent with stars with a fixed age and
metallicity, and varying mass.  The halo members show a roughly
consistent main sequence, but with somewhat greater scatter.  One
possible explanation for this scatter is that the halo is more
contaminated by field stars.  Another possible explanation is
differential reddening.  The halo is reported to span 20$^\circ$
on-sky, and varies in position from about $b=-12^\circ$ to
$b=-20^\circ$, with the stars closest to the galactic plane also being
further from the Sun by up to 200\,pc (Figure~\ref{fig:gaia6d}).
Empirical evidence for differential reddening is discussed further in
Appendix~\ref{app:diffred}. A priori, both field star contamination
and differential reddening could well play a role.

@@

% TODO: fix the fact that the figure currently only shows Randich+18
Figure~\ref{fig:lithium} shows the results.  At fixed stellar mass
(and age), the rapid rotators tend to show elevated lithium equivalent
widths. This effect is mostly apparent in the K dwarfs.  Similar
trends in the Pleiades were noted nearly three decades ago by
\citet{soderblom_evolution_1993}.  More recent studies have found
comparable results in the Pleiades, the Psc-Eri stream, and M\,35
\citep{bouvier_pleiades_lirot_2018,arancibia_2020,jeffries_m35_li_2020}.
The fact that we see the same trend across the core and halo of \cn\ 
{\it a)} supports the conclusion that the halo is coeval with the
core, and {\it b)} suggests that the lithium-rotation correlation is
not explained by environmental effects such as the density of the cluster
during its embedded phase, but is instead tied to physical processes
acting on all low-mass pre-main sequence stars.

@@

The authors thank X and Y for fruitful discussions.
%
L.G.B. and J.H. acknowledge support by the TESS GI Program, program
NUMBER, through NASA grant NUMBER.
%
This study was based in part on observations at Cerro Tololo
Inter-American Observatory at NSF's NOIRLab (NOIRLab Prop. ID
2020A-0146; 2020B-NUMBER PI: L{.}~Bouma), which is managed by the
Association of Universities for Research in Astronomy (AURA) under a
cooperative agreement with the National Science Foundation.
%
ACKNOWLEDGE PFS / CAMPANAS.
%
This paper includes data collected by the TESS mission, which are
publicly available from the Mikulski Archive for Space Telescopes
(MAST).
%
Funding for the TESS mission is provided by NASA's Science Mission
directorate.
%
We thank the TESS Architects (George Ricker, Roland Vanderspek, Dave
Latham, Sara Seager, Josh Winn, Jon Jenkins) and the many TESS team
members for their efforts to make the mission a continued success.
%

@@

%
% TODO: the following two paragraphs, when quoted, are ENTIRELY FROM
% the relevant papers. they should probably be nixed in favor of
% concision.
%
\citetalias{cantatgaudin_gaia_2018} applied a procedure that 
yielded what we call ``the core''.  Their procedure was to first query
a Gaia DR2 cone around the previously reported RA and dec of the
cluster, and within $\pm 0.5$\,mas of its previously reported
distance. 
% TODO: which was it?
The outer radius of their cone was either r2 from MWSC (CITE
Kharchenko 2013), or twice the ``cluster radius'' listed by DAML/Dias (CITE).  No
proper motion cut was applied.  They then applied an unsupervised
classification scheme called UPMASK to $G<18$\,mag stars within this
cone (CITE).  The steps of UPMASK are first to perform a k-means
clustering in the ``astrometric space'' ($\mu_{\alpha'}, \mu_\delta,
\pi$).  Then, a ``veto'' step is applied to assess whether the groups
of stars output from the k-means clustering are or are not more
concentrated than a random distribution.  This is implemented by
``comparing the total branch length of the minimum spanning tree
connecting the stars with the expected branch length for a random
uniform distribution covering the investigated field of view''.  ``To
turn this yes/no flag to a membership probability,
\citet{cantatgaudin_gaia_2018} then redraw new values of
($\mu_{\alpha'}, \mu_\delta, \pi$) for each source based on the listed
value, uncertainty, and covariance.  After a certain number of
redrawings, the final probability is the frequency with which a given
star passes the veto".  In the case of \cn, this yielded a reported
``\texttt{r50}'' within which half of the cluster members were found
to be within 0.496$^\circ$.
When we selected candidate \cn\ members from the results of
\citetalias{cantatgaudin_gaia_2018}, we opted to include all candidate members
with reported membership probability exceeding 10\%.
While this seems {\it a priori} low, our results ({\bf SECTION XXX})
show that this ``membership probabilty'' severely underestimates the
purity of the \citetalias{cantatgaudin_gaia_2018} sample for \cn. 
%TODO: quantify
Their false positive rate across the sample is more like 1-5\%.

@@

@@

To provide a basis for comparison, we also opted to search the
``calibration'' light curves ($Rp<13$) that were created as a part of
the CDIPS project.  Over the southern sky (Sectors 1-13 of TESS), this
corresponded to a sample of \ncalibration stars.  Cross-matching these
against the \nnbhd randomly drawn stars in the neighborhood of \cn yielded
\nnbhdcalibstar unique stars, with a cumulative total of \nnbhdcaliblc TESS
sectors observed.
The magnitude cut of $Rp<13$ at the distances of the neighborhood sample
corresponds to an extinction-corrected color cutoff of $(Bp-Rp)_0
\approx 0.80$, or spectral types of $\approx$G1V.
This reaches sufficiently far down the main sequence to enable a
comparison to the cluster star sample.

>>>>

As part of this broader project, we focus here on a modest question:
in just a single rich southern open cluster, is the cluster halo truly
coeval with the core?  We chose NGC\,2516 for this analysis since it
is young ($\approx$120\,Myr) and sufficiently nearby
($d\approx400$\,pc) to facilitate photometric rotation measurements
using TESS, and some degree of spectroscopic analysis.  Its halo is
decidedly non-spherical: \citetalias{kounkel_untangling_2019} reported
its dimensions as $\approx20\times10\times350$\,pc.  We want to know:
is the halo real? To what extent can we use Gaia alone to reliably
identify age-dated needles in the haystack of field stars?  What are
the implications, observationally and theoretically, if we can
identify the dispersed halos of open clusters?

@@

\section{A 300\,pc Halo around a Core?}
\label{sec:gaia}

@@

We selected candidate \cn\ members based on those available in the
literature.  While a number of pre-Gaia catalogs were available, the
purity and accuracy of the Gaia-derived results are the current state
of the art.
We therefore adopted what we viewed as the most interesting clustering
samples to compare: those of \citet{cantatgaudin_gaia_2018}
(\citetalias{cantatgaudin_gaia_2018}),
\citet{kounkel_untangling_2019}
(\citetalias{kounkel_untangling_2019}),
and \citet{meingast_2021} (\citetalias{meingast_2021}).
A useful visualization of these samples is available online\footnote{
  \url{https://homepage.univie.ac.at/stefan.meingast/coronae.html},
  made by \citet{meingast_2021}.  }.
While we could have performed our own clustering analysis based on the
Gaia data, such an effort would be only replicating the work of these
investigators.  We opt to instead use their studies as starting
points.

%FIXME FIXME FIXME FROM HERE

% Theia 613
\citetalias{cantatgaudin_gaia_2018} reported 1106 candidate members of
\cn, brighter than $G=18$\,mag.  \citet{kounkel_untangling_2019}
reported 3003 candidate members, and included stars up to
$\approx$1\,mag fainter.  
Finally, 1577 of the 1860 \citet{meingast_2021} sources (85\%)
were included in the \citet{kounkel_untangling_2019} sample.  
Each study applied unsupervised clustering techniques
to the second Gaia Data Release, as detailed in
Appendix~\ref{app:clustering}.

%FIXME 
\citetalias{cantatgaudin_gaia_2018} considered a cone of radius {\bf
XX DEGREES} centered on the cluster's position.
Given the classically reported tidal radius of {\it e.g.,} 9\,pc, CITE
Piskunov+2008, which is 8$^\circ$ at the cluster distance of
$\approx$400\,pc, perhaps they should have considered stars further
out.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
			\includegraphics[width=0.85\textwidth]{f2a.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf HR diagram of NGC\,2516.}
    The photometric data are from Gaia EDR3; no reddening correction
    has been applied.  The core (black) shows a main-sequence and
    turnoff consistent with stars of a fixed age and metallicity.  The
    halo (blue) is similar, though the binary sequence is less
    defined.  At fixed color, the faintest M dwarfs in the core and
    halo are brighter than in the field star comparison sample (gray),
    consistent with these stars having not yet reached the
    main-sequence.
    \label{fig:hr}
  }
\end{figure*}

The first check on the plausibility of the candidate cluster members
being coeval was already performed by
\citetalias{cantatgaudin_gaia_2018},
\citetalias{kounkel_untangling_2019}, and more recently by
\citet{meingast_2021}.  That check is to see whether the HR diagrams
of the cluster members support the claim that they are coeval.

Figure~\ref{fig:hr} presents similar results to what these
investigators have already found.  The core members of the cluster
show a clean sequence consistent with stars with a fixed age and
metallicity, and varying mass.  The halo members show a roughly
consistent main-sequence, but with somewhat greater scatter.  One
possible explanation for this scatter is that the halo is more
contaminated by field stars.  Another possible explanation is
differential reddening.  The halo is reported to span 20$^\circ$
on-sky, and varies in position from about $b=-12^\circ$ to
$b=-20^\circ$, with the stars closest to the galactic plane also being
further from the Sun by up to 200\,pc (Figure~\ref{fig:gaia6d}).
Empirical evidence for differential reddening is discussed further in
Appendix~\ref{app:diffred}. A priori, both field star contamination
and differential reddening could well play a role.

@@

% TODO: fix the fact that the figure currently only shows Randich+18
Figure~\ref{fig:lithium} shows the results.  At fixed stellar mass
(and age), the rapid rotators tend to show elevated lithium equivalent
widths. This effect is mostly apparent in the K dwarfs.  Similar
trends in the Pleiades were noted nearly three decades ago by
\citet{soderblom_evolution_1993}.  More recent studies have found
comparable results in the Pleiades, the Psc-Eri stream, and M\,35
\citep{bouvier_pleiades_lirot_2018,arancibia_2020,jeffries_m35_li_2020}.
The fact that we see the same trend across the core and halo of \cn\ 
{\it a)} supports the conclusion that the halo is coeval with the
core, and {\it b)} suggests that the lithium-rotation correlation is
not explained by environmental effects such as the density of the cluster
during its embedded phase, but is instead tied to physical processes
acting on all low-mass pre-main-sequence stars.

@@

The authors thank X and Y for fruitful discussions.
%
L.G.B. and J.H. acknowledge support by the TESS GI Program, program
NUMBER, through NASA grant NUMBER.
L.G.B. was also supported by a XXXX Fellowship from Princeton
University.
%
This study was based in part on observations at Cerro Tololo
Inter-American Observatory at NSF's NOIRLab (NOIRLab Prop. ID
2020A-0146; 2020B-NUMBER PI: L{.}~Bouma), which is managed by the
Association of Universities for Research in Astronomy (AURA) under a
cooperative agreement with the National Science Foundation.
%
ACKNOWLEDGE PFS / CAMPANAS.
%
This paper includes data collected by the TESS mission, which are
publicly available from the Mikulski Archive for Space Telescopes
(MAST).
%
Funding for the TESS mission is provided by NASA's Science Mission
directorate.
%
We thank the TESS Architects (George Ricker, Roland Vanderspek, Dave
Latham, Sara Seager, Josh Winn, Jon Jenkins) and the many TESS team
members for their efforts to make the mission a continued success.
%

@@

%
% TODO: the following two paragraphs, WHEN QUOTED, are ENTIRELY FROM
% the relevant papers. they should probably be nixed in favor of
% concision.
%
\citetalias{cantatgaudin_gaia_2018} applied a procedure that 
yielded what we call ``the core''.  Their procedure was to first query
a Gaia DR2 cone around the previously reported RA and dec of the
cluster, and within $\pm 0.5$\,mas of its previously reported
distance. 
% TODO: which was it?
The outer radius of their cone was either r2 from MWSC (CITE
Kharchenko 2013), or twice the ``cluster radius'' listed by DAML/Dias (CITE).  No
proper motion cut was applied.  They then applied an unsupervised
classification scheme called UPMASK to $G<18$\,mag stars within this
cone (CITE).  The steps of UPMASK are first to perform a k-means
clustering in the ``astrometric space'' ($\mu_{\alpha'}, \mu_\delta,
\pi$).  Then, a ``veto'' step is applied to assess whether the groups
of stars output from the k-means clustering are or are not more
concentrated than a random distribution.  This is implemented by
``comparing the total branch length of the minimum spanning tree
connecting the stars with the expected branch length for a random
uniform distribution covering the investigated field of view''.  ``To
turn this yes/no flag to a membership probability,
\citet{cantatgaudin_gaia_2018} then redraw new values of
($\mu_{\alpha'}, \mu_\delta, \pi$) for each source based on the listed
value, uncertainty, and covariance.  After a certain number of
redrawings, the final probability is the frequency with which a given
star passes the veto".  In the case of \cn, this yielded a reported
``\texttt{r50}'' within which half of the cluster members were found
to be within 0.496$^\circ$.
When we selected candidate \cn\ members from the results of
\citetalias{cantatgaudin_gaia_2018}, we opted to include all candidate members
with reported membership probability exceeding 10\%.
While this seems {\it a priori} low, our results ({\bf SECTION XXX})
show that this ``membership probabilty'' severely underestimates the
purity of the \citetalias{cantatgaudin_gaia_2018} sample for \cn. 
%TODO: quantify
Their false positive rate across the sample is more like 1-5\%.

@@

\citetalias{meingast_2021} did the same as
\citetalias{kounkel_untangling_2019}, but required A) the tangential
velocity dispersions to be smaller, and B) deconvolved the distances.
{\bf FIXME FIXME FIXME}
{\bf FIXME FIXME FIXME}
{\bf FIXME FIXME FIXME}

@@

To provide a basis for comparison, we also opted to search the
``calibration'' light curves ($Rp<13$) that were created as a part of
the CDIPS project.  Over the southern sky (Sectors 1-13 of TESS), this
corresponded to a sample of \ncalibration stars.  Cross-matching these
against the \nnbhd randomly drawn stars in the neighborhood of \cn yielded
\nnbhdcalibstar unique stars, with a cumulative total of \nnbhdcaliblc TESS
sectors observed.
The magnitude cut of $Rp<13$ at the distances of the neighborhood sample
corresponds to an extinction-corrected color cutoff of $(Bp-Rp)_0
\approx 0.80$, or spectral types of $\approx$G1V.
This reaches sufficiently far down the main-sequence to enable a
comparison to the cluster star sample.
